---
title: "Accessing EuPathDB Resources using AnnotationHub"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Accessing EuPathDB Resources using AnnotationHub}
  %\VignetteEngine{knitr::rmarkdown}
  % \VignetteKeyword{eupathdb, annotations}
  \usepackage[utf8]{inputenc}
---

```{r style, echo=FALSE, results='asis', message=FALSE}
BiocStyle::markdown()
```

**Authors**: [V. Keith Hughitt](mailto:keith.hughitt@nih.gov)<br />
**Modified:** `r file.info("EuPathDB.Rmd")$mtime`<br />
**Compiled**: `r date()`

# Overview

This tutorial describes how to query and make use of annotations retrieved from
[EuPathDB : The Eukaryotic Pathogen Genomics Resource](http://eupathdb.org/eupathdb/)
using [AnnotationHub](http://bioconductor.org/packages/release/bioc/html/AnnotationHub.html).

For more information on using AnnotationHub, check out the AnnotationHub
vignettes:

- [AnnotationHub: Access the AnnotationHub Web Service](http://bioconductor.org/packages/release/bioc/vignettes/AnnotationHub/inst/doc/AnnotationHub-HOWTO.html)
- [AnnotationHub How-To’s](http://bioconductor.org/packages/release/bioc/vignettes/AnnotationHub/inst/doc/AnnotationHub-HOWTO.html)

The resources described in this tutorial were generating using GFF files and
web API requests made to the various EuPathDB databases (TriTrypDB, ToxoDB,
etc.) Only organisms with annotated genomes (those for which GFF files are
available) are accessible through AnnotationHub.

The two main resources provided are:

- [OrgDb](https://www.bioconductor.org/help/workflows/annotation/annotation/#OrgDb)
- [GRanges](http://bioconductor.org/packages/release/bioc/html/GenomicRanges.html)

OrgDB objects for an organism include basic gene-level information such as:

- Gene ID
- Gene description
- Chromosome number
- GO terms assocated with gene
- KEGG Pathways associated with gene
- Etc.

For some organisms, [InterPro](https://www.ebi.ac.uk/interpro/) protein domain
information is also available (in some cases, however, even though InterPro domain
information is available through EuPathDB, it is too large to be included in
the current AnnotationHub resources).

For more information about working with Bioconductor annotation resources, see:

- [Genomic Annotation Resources in Bioconductor ](https://www.bioconductor.org/help/workflows/annotation/annotation/)

# Installation

If you don't already have AnnotationHub installed on your system, use
`biocLite` to install the package:

```{r, eval=FALSE}
source("http://bioconductor.org/biocLite.R")
biocLite('AnnotationHub')
```

# Getting started

To begin, let's create a new `AnnotationHub` connection and use it to query
AnnotationHub for all EuPathDB resources.

```{r}
library('AnnotationHub')

# create an AnnotationHub connection
ah = AnnotationHub()

# search for all EuPathDB resources
meta = query(ah, "EuPathDB")

length(meta)
head(meta)

# types of EuPathDB data available
table(meta$rdataclass)

# distribution of resources by specific databases
table(meta$dataprovider)

# List of organisms for which resources are available
length(unique(meta$species))
head(unique(meta$species))
```

# Working with EuPathDB OrgDB resources

Next, we will see how you can query AnnotationHub for EuPathDB OrgDB resources.

To begin, create an AnnotationHub connection, if you have not already done so,
as shown in the section above.

Next, use the `query` function to search for your organism of interest, and
store the result.

```{r}
res <- query(ah, 'Leishmania major strain Friedlin', 'OrgDb')
res
```

The result includes a single record, "AH56967". The record can be accessed from
the result variable using list-like indexing:

```{r}
orgdb <- res[['AH56967']]
class(orgdb)
```

We can see that we now have an OrgDB instance, and as such, we can use the
usual methods available for working this OrgDB objects, including:

- `columns()`
- `keys()`
- `select()`


```{r}
columns(orgdb)

gids <- keys(orgdb, keytype='GID')
head(gids)

dat <- select(orgdb, keys=gids, keytype='GID', columns=c('TYPE', 'CHR'))

head(dat)

table(dat$TYPE)
table(dat$CHR)
```

# Working with EuPathDB GRanges resources

```{r}
res
```
