---
title: "What happens when I invoke makeAnnotationHubMetadata()?"
author: "atb abelew@gmail.com"
date: "`r Sys.Date()`"
output:
  html_document:
    code_download: true
    code_folding: show
    fig_caption: true
    fig_height: 7
    fig_width: 7
    highlight: tango
    keep_md: false
    mode: selfcontained
    number_sections: true
    self_contained: true
    theme: readable
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
  rmdformats::readthedown:
    code_download: true
    code_folding: show
    df_print: paged
    fig_caption: true
    fig_height: 7
    fig_width: 7
    highlight: tango
    width: 300
    keep_md: false
    mode: selfcontained
    toc_float: true
  BiocStyle::html_document:
    code_download: true
    code_folding: show
    fig_caption: true
    fig_height: 7
    fig_width: 7
    highlight: tango
    keep_md: false
    mode: selfcontained
    toc_float: true
---

<style type="text/css">
body, td {
  font-size: 16px;
}
code.r{
  font-size: 16px;
}
pre {
 font-size: 16px
}
</style>

```{r options, include=FALSE}
library("hpgltools")
tt <- devtools::load_all("/data/hpgltools")
knitr::opts_knit$set(width=120,
                     progress=TRUE,
                     verbose=TRUE,
                     echo=TRUE)
knitr::opts_chunk$set(error=TRUE,
                      dpi=96)
old_options <- options(digits=4,
                       stringsAsFactors=FALSE,
                       knitr.duplicate.label="allow")
ggplot2::theme_set(ggplot2::theme_bw(base_size=10))
rundate <- format(Sys.Date(), format="%Y%m%d")
ver <- "20180822"

rmd_file <- "make_data.Rmd"
```

# Creating AnnotationHub Metadata!

In this r markdown file, I am copying my make-data.R script into a R markdown
document so that I can attempt to prove in a reproducible fashion that the csv
and metadata I generate are valid.

I am removing the portion of my script which actually generated the _data_,
because I do not want to wait for it to finish checking all the packages.

## Loading my configuration

I am currently 'parallelizing' this script by invoking a copy of it for each
eupathdb service. When each service completes, I come back and run make-data.R,
which checks each data file, tries to regenerate anything which failed, and is
responsible for making the final metadata.  As a result, I use a single
configuration file to define the version of the eupathdb to download, etc.
config.R also loads my copy of the EuPathDB package.  testthat is used when
generating the actual data so that I can check each package.

```{r configuration}
library(testthat)
source("config.R")
webservice <- "eupathdb"

eupathdb_version
bioc_version
```

## Download the metadata from the various eupathdb web services.

In this block I copy/paste the metadata downloader.

```{r download_metadata}
meta <- download_eupathdb_metadata(
  bioc_version=bioc_version, build_dir=build_dir, overwrite=TRUE, webservice=webservice,
  verbose=TRUE, eupathdb_version=eupathdb_version, write_csv=TRUE)
all_metadata <- meta[["valid"]]
write_eupathdb_metadata(all_metadata, bioc_version=bioc_version, eupathdb_version=eupathdb_version)
end <- nrow(all_metadata)

summary(all_metadata)
end
```

The eupathdb provided me apparently 376 entries that my script will then attempt
to make the various packages for.

## The package creation body

The following block will not be run, but is what actually does the work of
making the various rda/sqlite databases.

```{r do_work, eval=FALSE}
start <- 1
for (it in start:end) {
  entry <- all_metadata[it, ]
  species <- entry[["Species"]]
  message("Starting generation of ", species, ", which is ", it, " of ", end, " species.")
  pkgnames <- get_eupathdb_pkgnames(entry)
  if (isTRUE(bsgenome)) {
    bsgenome_result <- make_eupathdb_bsgenome(entry, workdir = build_dir, eupathdb_version=eupathdb_version, copy_s3=TRUE)
    expected <- "bsgenome_name"
    actual <- names(bsgenome_result)
    testthat::test_that("Does make_eupathdb_bsgenome return something sensible?", {
      expect_equal(expected, actual)
    })
    results[["bsgenome"]][[species]] <- bsgenome_result
  }
  if (isTRUE(orgdb)) {
    orgdb_result <- make_eupathdb_orgdb(entry, workdir = build_dir, eupathdb_version=eupathdb_version, copy_s3=TRUE)
    if (is.null(orgdb_result)) {
      message("There is insufficient data for ", species, " to make the OrgDB.")
    } else {
      actual <- orgdb_result[["orgdb_name"]]
      expected <- pkgnames[["orgdb"]]
      test_that("Does make_eupathdb_orgdb return something sensible?", {
        expect_equal(expected, actual)
      })
      results[["orgdb"]][[species]] <- orgdb_result
    }
  }
  if (isTRUE(txdb)) {
    txdb_result <- make_eupathdb_txdb(entry, workdir = build_dir, eupathdb_version=eupathdb_version, copy_s3=TRUE)
    if (is.null(txdb_result)) {
      message("Unable to create the txdb package.")
    } else {
      actual <- txdb_result[["txdb_name"]]
      expected <- pkgnames[["txdb"]]
      test_that("Does make_eupathdb_txdb return something sensible?", {
        expect_equal(expected, actual)
      })
      results[["txdb"]][[species]] <- txdb_result[["txdb_name"]]
    }
  }
  if (isTRUE(organismdb)) {
    organ_result <- make_eupathdb_organismdbi(entry, workdir = build_dir, eupathdb_version=eupathdb_version, copy_s3=TRUE)
    actual <- organ_result[["organismdb_name"]]
    expected <- pkgnames[["organismdbi"]]
    test_that("Does make_eupathdb_organismdbi return something sensible?", {
      expect_equal(expected, actual)
    })
    results[["organismdbi"]] <- organ_result
  }
} ## End iterating over every entry in the eupathdb metadata.
```

## The final checks

Once the data has been generated (or not), the final task is to check the
metadata and the generated files.  The following block accomplishes this:

the variables bsgenome and organismdb are of course, FALSE in this invocation.

### Check the orgdb metadata

One might wonder while I do a file.copy() here.  I actually run all this stuff
in a separate tree in order to keep it separate from my EuPathDB codebase,
mostly because I need to run it on our computing cluster where there is a
scratch disk of sufficient size for these files.

```{r final_orgdb}
if (isTRUE(orgdb)) {
  org_csv <- check_csv(build_dir, file_type="OrgDb", bioc_version=bioc_version, eupathdb_version=eupathdb_version)
  org_files <- check_files("OrgDb", bioc_version=bioc_version, eupathdb_version=eupathdb_version)
  csv_copy_path <- file.path(path.package("EuPathDB"), "inst", "extdata", org_csv)
  copied <- file.copy(org_csv, csv_copy_path, overwrite=TRUE)
  expect_true(copied)
  org_checked <- AnnotationHubData::makeAnnotationHubMetadata(path.package("EuPathDB"), org_csv)
  save(list=c("org_checked"), file=file.path(build_dir, "orgdb_metadata.rda"))
  class(org_checked)
  summary(org_checked)
}
```

### Check the txdb metadata

```{r final_txdb}
if (isTRUE(txdb)) {
  txdb_csv <- check_csv(build_dir, file_type="TxDb", bioc_version=bioc_version, eupathdb_version=eupathdb_version)
  tx_files <- check_files("TxDb", bioc_version=bioc_version, eupathdb_version=eupathdb_version)
  csv_copy_path <- file.path(path.package("EuPathDB"), "inst", "extdata", txdb_csv)
  copied <- file.copy(txdb_csv, csv_copy_path, overwrite=TRUE)
  expect_true(copied)
  tx_checked <- AnnotationHubData::makeAnnotationHubMetadata(path.package("EuPathDB"), txdb_csv)
  save(list=c("tx_checked"), file=file.path(build_dir, "txdb_metadata.rda"))
  class(tx_checked)
  summary(tx_checked)
}
```

### Check the granges

```{r final_granges}
if (isTRUE(granges)) {
  grange_csv <- check_csv(build_dir, file_type="GRanges", bioc_version=bioc_version, eupathdb_version=eupathdb_version)
  grange_files <- check_files("GRanges", bioc_version=bioc_version, eupathdb_version=eupathdb_version)
  csv_copy_path <- file.path(path.package("EuPathDB"), "inst", "extdata", grange_csv)
  copied <- file.copy(grange_csv, csv_copy_path, overwrite=TRUE)
  expect_true(copied)
  grange_checked <- AnnotationHubData::makeAnnotationHubMetadata(path.package("EuPathDB"), grange_csv)
  save(list=c("grange_checked"), file=file.path(build_dir, "granges_metadata.rda"))
  class(grange_checked)
  summary(grange_checked)
}
```

# Full script

For full disclosure, here is the full text of my script.  In the above,
I am just invoking the relevant portions for this discussion.

```{r full_script, eval=FALSE}
#!/usr/bin/env Rscript
library(testthat)
source("config.R")
webservice <- "eupathdb"
meta <- download_eupathdb_metadata(
  bioc_version=bioc_version, overwrite=TRUE, webservice=webservice,
  verbose=TRUE, eupathdb_version=eupathdb_version, write_csv=TRUE)
all_metadata <- meta[["valid"]]
write_eupathdb_metadata(all_metadata, bioc_version=bioc_version, eupathdb_version=eupathdb_version)
end <- nrow(all_metadata)

start <- 1
for (it in start:end) {
  entry <- all_metadata[it, ]
  species <- entry[["Species"]]
  message("Starting generation of ", species, ", which is ", it, " of ", end, " species.")
  pkgnames <- get_eupathdb_pkgnames(entry)
  if (isTRUE(bsgenome)) {
    bsgenome_result <- make_eupathdb_bsgenome(entry, workdir = build_dir, eupathdb_version=eupathdb_version, copy_s3=TRUE)
    expected <- "bsgenome_name"
    actual <- names(bsgenome_result)
    testthat::test_that("Does make_eupathdb_bsgenome return something sensible?", {
      expect_equal(expected, actual)
    })
    results[["bsgenome"]][[species]] <- bsgenome_result
  }
  if (isTRUE(orgdb)) {
    orgdb_result <- make_eupathdb_orgdb(entry, workdir = build_dir, eupathdb_version=eupathdb_version, copy_s3=TRUE)
    if (is.null(orgdb_result)) {
      message("There is insufficient data for ", species, " to make the OrgDB.")
    } else {
      actual <- orgdb_result[["orgdb_name"]]
      expected <- pkgnames[["orgdb"]]
      test_that("Does make_eupathdb_orgdb return something sensible?", {
        expect_equal(expected, actual)
      })
      results[["orgdb"]][[species]] <- orgdb_result
    }
  }
  if (isTRUE(txdb)) {
    txdb_result <- make_eupathdb_txdb(entry, workdir = build_dir, eupathdb_version=eupathdb_version, copy_s3=TRUE)
    if (is.null(txdb_result)) {
      message("Unable to create the txdb package.")
    } else {
      actual <- txdb_result[["txdb_name"]]
      expected <- pkgnames[["txdb"]]
      test_that("Does make_eupathdb_txdb return something sensible?", {
        expect_equal(expected, actual)
      })
      results[["txdb"]][[species]] <- txdb_result[["txdb_name"]]
    }
  }
  if (isTRUE(organismdb)) {
    organ_result <- make_eupathdb_organismdbi(entry, workdir = build_dir, eupathdb_version=eupathdb_version, copy_s3=TRUE)
    actual <- organ_result[["organismdb_name"]]
    expected <- pkgnames[["organismdbi"]]
    test_that("Does make_eupathdb_organismdbi return something sensible?", {
      expect_equal(expected, actual)
    })
    results[["organismdbi"]] <- organ_result
  }
} ## End iterating over every entry in the eupathdb metadata.

if (isTRUE(bsgenome)) {
  bs_csv <- check_csv(build_dir, file_type="BSgenome", bioc_version=bioc_version, eupathdb_version=eupathdb_version)
  bs_files <- check_files("BSgenome", bioc_version=bioc_version, eupathdb_version=eupathdb_version)
  csv_copy_path <- file.path(path.package("EuPathDB"), "inst", "extdata", bs_csv)
  file.copy(bs_csv, csv_copy_path)
  bs_checked <- AnnotationHubData::makeAnnotationHubMetadata(path.package("EuPathDB"), bs_csv)
  save(list=c("bs_checked"), file=file.path(build_dir, "bsgenome_metadata.rda"))
  bs_checked[[1]]
}
if (isTRUE(orgdb)) {
  org_csv <- check_csv(build_dir, file_type="OrgDb", bioc_version=bioc_version, eupathdb_version=eupathdb_version)
  org_files <- check_files("OrgDb", bioc_version=bioc_version, eupathdb_version=eupathdb_version)
  csv_copy_path <- file.path(path.package("EuPathDB"), "inst", "extdata", org_csv)
  file.copy(org_csv, csv_copy_path)
  org_checked <- AnnotationHubData::makeAnnotationHubMetadata(path.package("EuPathDB"), orgdb_csv)
  save(list=c("org_checked"), file=file.path(build_dir, "orgdb_metadata.rda"))
  org_checked[[1]]
}
if (isTRUE(txdb)) {
  txdb_csv <- check_csv(build_dir, file_type="TxDb", bioc_version=bioc_version, eupathdb_version=eupathdb_version)
  tx_files <- check_files("TxDb", bioc_version=bioc_version, eupathdb_version=eupathdb_version)
  csv_copy_path <- file.path(path.package("EuPathDB"), "inst", "extdata", txdb_csv)
  file.copy(txdb_csv, csv_copy_path)
  tx_checked <- AnnotationHubData::makeAnnotationHubMetadata(path.package("EuPathDB"), txdb_csv)
  save(list=c("tx_checked"), file=file.path(build_dir, "txdb_metadata.rda"))
  tx_checked[[1]]
}
if (isTRUE(organismdb)) {
  organ_csv <- check_csv(build_dir, file_type="OrganismDbi", bioc_version=bioc_version, eupathdb_version=eupathdb_version)
  organ_files <- check_files("OrganismDbi", bioc_version=bioc_version, eupathdb_version=eupathdb_version)
  csv_copy_path <- file.path(path.package("EuPathDB"), "inst", "extdata", organ_csv)
  file.copy(organ_csv, csv_copy_path)
  organ_checked <- AnnotationHubData::makeAnnotationHubMetadata(path.package("EuPathDB"), organ_csv)
  save(list=c("organ_checked"), file=file.path(build_dir, "organismdbi_metadata.rda"))
  organ_checked[[1]]
}
if (isTRUE(granges)) {
  grange_csv <- check_csv(build_dir, file_type="GRanges", bioc_version=bioc_version, eupathdb_version=eupathdb_version)
  grange_files <- check_files("GRanges", bioc_version=bioc_version, eupathdb_version=eupathdb_version)
  csv_copy_path <- file.path(path.package("EuPathDB"), "inst", "extdata", grange_csv)
  file.copy(grange_csv, csv_copy_path)
  grange_checked <- AnnotationHubData::makeAnnotationHubMetadata(path.package("EuPathDB"), grange_csv)
  save(list=c("grange_checked"), file=file.path(build_dir, "granges_metadata.rda"))
  grange_checked[[1]]
}
```




```{r saveme}
pander::pander(sessionInfo())
message(paste0("This is hpgltools commit: ", get_git_commit()))
this_save <- paste0(gsub("\\.Rmd", "", rmd_file), "-v", ver, ".rda.xz")
message(paste0("Saving to ", this_save))
tmp <- sm(saveme(filename=this_save))
```
